Classless Java

### 2016-03-23 OOPSLA16/sections/Introduction.tex
revised S1: typos, grammar errors

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

This paper presents a third alternative: the concept of
\textbf{Interface based} (IB) object oriented languages, where objects
implement interfaces directly. In IB interfaces own the implementation
for the behaviour, which is structurally defined in their
interface. Programmers do not define objects directly, but delegate
the task to \emph{object interfaces}, whose role is similar to non
abstract classes in CB languages. Instantiation of objects is
done using static factory methods in object interfaces.

A key challenge in IB programming lies in how to model state, which is
fundamental to have stateful objects. All abstract operations in an
object interface are interpreted as \emph{abstract state
  operations}. The abstract state operations include various common
utility methods (such as getters and setters, or clone-like
methods). Objects are the only responsible to define the ultimate
behaviour of a method and, for example, if such method is just a
setter. Anything related to state is completely contained in the
instances and does not leak into the inheritance logic.  In CB
programming, the structure of the state is fixed and can be only grown
by inheritance.  In contrast, in IB programming the state is never
fixed, and methods that ``looks like'' abstract setters and getters
can always receive an explicit implementation down in the inheritance
chain, improving \textbf{modularity and flexibility}.  That is, the
concept of abstract state is more fluid.

@@

Finally, another advantage of IB programming is its support for
multiple inheritance. The literature provides good examples on how
easy and modular it is to combine multiple sources of pure behaviour,
using mechanisms such as traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and in Java8 default methods~\cite{goetz12fdefenders} bring some of the
advantages of traits into Java. In contrast, the literature~\cite{Sak89dis,bracha90mixin,malayeri2009cz}
is also rich on how hard it is to modularly combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB programming there is only multiple
interface inheritance, yet programmers can still use state via the
abstract state operations. IB programming enables powerful 
idioms using multiple inheritance and state.

@@

IB could be explained by defining a novel language, with new syntax
and semantics. However, this would have a steep learning curve.  We
take a different approach instead. For the sake of providing a more
accessible explanation, we will embed our ideas directly into Java. 
Our IB embedding relies on the
new features of Java 8 for interfaces: interface \emph{static methods}; and
\emph{default methods}, which allow interfaces to have method
implementations. In the context of Java, what we propose is a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Classless Java}.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers.  The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
As a result, we experiment object interfaces with several interesting
Java programs, and conduct various case studies.  Finally, we also
formally define the behaviour of our \Q|Obj| annotation and discuss
its properties.

@@

To evaluate the usefulness of object interfaces, we illustrate
\numOfCaseStudies \bruno{needs updates}\yanlin{updated} 
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The third
application is a case study for a simple Maze game implemented with 
multiple inheritance. For this application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. The last
application is a relatively larger case study of refactoring of a simple interpreter, showing that our
approach can benefit modularity and scale to real code base. Noteworthy, the first three applications are implemented 
without defining a single class! The last one contains several classes that cannot be 
converted to interfaces due to the limitation of our implementation of \mixin.

>>>>

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects are all you have).
In class based languages an object is an instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

This paper presents a third alternative: the concept of
\textbf{Interface based} (IB) object oriented languages, where objects
implement interfaces directly. In IB interfaces own the implementation
for the behaviour, which is structurally defined in their
interface. Programmers do not define objects directly, but delegate
the task to \emph{object interfaces}, whose role is similar to non-abstract 
classes in CB languages. Instantiation of objects is
done using static factory methods in object interfaces.

A key challenge in IB programming lies in how to model state, which is
fundamental to have stateful objects. All abstract operations in an
object interface are interpreted as \emph{abstract state
  operations}. The abstract state operations include various common
utility methods (such as getters and setters, or clone-like
methods). Objects are only responsible to define the ultimate
behaviour of a method and, for example, if such a method is just a
setter. Anything related to state is completely contained in the
instances and does not leak into the inheritance logic.  In CB
programming, the structure of the state is fixed and can be only grown
by inheritance.  In contrast, in IB programming the state is never
fixed, and methods that look like abstract setters and getters
can always receive an explicit implementation down in the inheritance
chain, improving \textbf{modularity and flexibility}.  That is, the
concept of abstract state is more fluid.

@@

Finally, another advantage of IB programming is its support for
multiple inheritance. The literature provides good examples on how
easy and modular it is to combine multiple sources of pure behaviour,
using mechanisms such as traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and in Java 8 default methods~\cite{goetz12fdefenders} bring some of the
advantages of traits into Java. In contrast, the literature~\cite{Sak89dis,bracha90mixin,malayeri2009cz}
is also rich on how hard it is to modularly combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB programming there is only multiple
interface inheritance, yet programmers can still use state via the
abstract state operations. IB programming enables powerful 
idioms using multiple inheritance and state.

@@

IB could be explained by defining a novel language, with new syntax
and semantics. However, this would have a steep learning curve.  We
take a different approach instead. For the sake of providing a more
accessible explanation, we will embed our ideas directly into Java. 
Our IB embedding relies on the
new features of Java 8 for interfaces: interface \emph{static methods} and
\emph{default methods}, which allow interfaces to have method
implementations. In the context of Java, what we propose is a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Classless Java}.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java 8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers.  The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
As a result, we experiment object interfaces with several interesting
Java programs, and conduct various case studies.  Finally, we also
formally define the behaviour of our \mixin annotation and discuss
its properties.

@@

To evaluate the usefulness of object interfaces, we illustrate
\numOfCaseStudies \bruno{needs updates}\yanlin{updated} 
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The third
application is a case study for a simple Maze game implemented with 
multiple inheritance. For this application we show that there is a
significant reduction in the number of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. The last
application is a relatively larger case study of refactoring of a simple interpreter, showing that our
approach can benefit modularity and scale to real code base. Noteworthy is the fact that the first three applications are implemented 
without defining a single class! The last one contains several classes that cannot be 
converted to interfaces due to the limitation of our implementation of \mixin.
